---
title: "Supplement 9 - diveMove calibration example"
date: "2024-09-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

A simple calibration example showing that 'offset' in calibrateDepth is to 'correct the surface value'. 
'offset' should not be used to ignore shallow dives (use 'dive.thr' for that purpose). Cases in the published literature have shown 
these commands are sometimes confused.


```{r load packages, echo=TRUE, results='hide', message=FALSE, warning=FALSE}

library(tidyverse)
library(diveMove)
library(here)
```

```{r code, warning = FALSE}
# Import TDR dive data 
dives <-  readRDS(here("data", "diveMove_example.rds"))
# Count how many dives are deeper than 5 m — the answer is 8.
plot(dives$date.time, dives$depth, pch = ".", col = "navy", cex = 2)  # plot the data
abline(h = 5, lty = 2) # add line at 5 m depth
abline(h = 30, lty = 2) # add line at 30 m depth

# Zoom in around 0 m — there is no drift, so no correction is needed for this TDR data.
plot(dives$date.time, dives$depth, pch = ".", ylim = c(-0.5, 2), col = "navy", cex = 2)
abline(h = -0.01)  # add line at -0.01 m depth

# No need to zero-offset correct this dive data

#now start creating the TDR dive object
filename = "Dive data"

tdr <- createTDR(time = dives$date.time, 
                 depth = dives$depth, 
                 speed = FALSE, 
                 dtime = 1,   #  sampling interval used in seconds
                 file = filename)

show(tdr)

#---------------------------------------
# The calibrateDepth step is important
#---------------------------------------
# Can use the filter method: this code ignore all dives shallower than 5 m.
# Correctly identifies 8 dives
tdr.calib <- calibrateDepth(tdr,
                            dive.thr = 5,       # only select dives deeper than 5 m.
                            zoc.method='filter',
                            k=c(3, 5760),
                            probs=c(0.5, 0.02),
                            #dry.thr=3600, wet.thr=30, interp.wet=FALSE, 
                            smooth.par=0.1,
                            knot.factor=20, 
                            descent.crit.q=0.01, ascent.crit.q=0.01,
                            na.rm=T)

# Be careful with the offset method. The following code is not correct.
# zoc.method="offset" does not determine which dives to ignore — it asks,
# "How wrong is the surface layer, and how much should I adjust the surface value?"

# Using offset=5 with the default dive threshold (4m) ignores all dives shallower than 9m!
tdr.calib_offset <- calibrateDepth(tdr,
                                   zoc.method="offset",
                                   offset=5,
                                   # dive.thr = 4,  # default
                                   descent.crit.q=0.01, ascent.crit.q=0.01,
                                   knot.factor=60)

# Only 7 dives are identified in this example, even though there are 8 dives deeper than 5 m.

# create dive summary metrics for each dive
tdr.dat = as.data.frame(tdr.calib@tdr)
dive.stats <- diveStats(tdr.calib)        

# create dive summary metrics for each dive 
tdr.dat_offset = as.data.frame(tdr.calib_offset@tdr)
dive.stats_offset <- diveStats(tdr.calib_offset)        

# plot the max dive depths from the filter method
plot(dive.stats$maxdep, ylim = c(0,38), pch = 15, xlab = "Dive number", ylab = "Maximum depth")  
abline(h = 30, lty = 2) # Note that all maximum depths are above 30 m, consistent with the data.
abline(h = 5, lty = 2) # add line at 5 m depth

# now add max dive depth from the offset method
# the max depths are under-estimated
# Dive 2 was not recognized, because is it shallower than 5 + 4  = 9 meter.
# (offset = 5 m and dive.thr = 4 m in calibrateDepth)
points(c(1, 3:8), dive.stats_offset$maxdep,  col = "blue", pch = 16)  # max dive depths

```

When 'offset=5' is incorrectly used, there is a systematic bias in maximum depth (blue circles in the plot above), and some dives deeper than the target threshold depth are missed. 

